---
title:  "A3_Process.Rmd / Heat-Related Mortality / Exposure Response"
author: "Hans-Aloys Wischmann, Institute for Public Health, Charité - Universitätsmedizin Berlin"
date:   "Written: 2023-04-12 to 2023-08-31 / Totally revised: 2024-11-07 to 2025-07-03 / Executed: `r Sys.Date()`"
output:
  officedown::rdocx_document:
    reference_docx: A4_Template.docx
    tables:
      style: Normal
    plots:
      style: Normal
---

```{r setup, include = FALSE}
  source("./A1_Parameters.R")
  library_wrapper(c("broom", "mgcv"))
```

# Prerequisites and Expected Execution Time

Prerequisites: A2_Preprocess.Rmd successfully completed

Expected execution time: About 7 minutes (12/24 physical/logical cores @ 5.7 GHz, 32 GB RAM)

# Pass 1 - Individual models by region, by age, by sex, by year: Onset of heat-related risk

Individual GAMs are computed for the weekly death rates (assuming a Gaussian error distribution) as a function of the mean temperature across the prior 0..3 days, separately by region(geo), by age groups, by sex, and by year. The onset (t_zero) for positive heat-related mortality risk is determined for each model (along with the slope (s_zero) at this point),  aggregated across all available years, and subtracted from the actual, observed temperatures, by region(geo), by age, by sex. Many individual models do not show a minimum (t_min), so that the computed estimates for the minima are less robust and can thus not serve as the temperature reference. As the data frame for each model contains only 26 data points (LAST_SUMMER_WEEK - FIRST_SUMMER_WEEK + 1), the seasonal terms are modeled using a smooth function with only 5 DoF. Data from two COVID years (2020 and 2022) are excluded to avoid potential (differential) confounding by pandemic mortality effects during summer.

```{r fit_individual_GAMs}
  # reload summer data file and geography data
  load(file = "../Processed/DeathsPopTemp.Rdata")
  load(file = "../Processed/Geography.Rdata")

  # document number of deaths processed
  n_total <- sum((df_summer %>% filter(!year %in% COVID_YEARS))$n_deaths)
  df_summer %>% mutate(geo = gsub("UK", "EN.WA", str_sub(geo, 1, 2))) %>%
    group_by(geo, year) %>% reframe(n_deaths = round(sum(n_deaths) * 0.001, 0)) %>% mutate(year = factor(year)) %>%
    pivot_wider(names_from = geo, values_from = n_deaths) %>%
    flextable() %>% set_caption(sprintf("Deaths processed [1000s] (excl. COVID years), Total = %d", n_total)) %>% autofit()

  # list of models to fit: by region, age group, sex, and year (1st pass, excluding years with COVID) or by age group and sex (2nd to 4th pass)
  m_age_sex          <- df_summer %>% select(              age, sex      ) %>% distinct()
  m_geo_age_sex_year <- df_summer %>% select(country, geo, age, sex, year) %>% distinct() %>% filter(!year %in% COVID_YEARS)

  # fit separate models by region, age group, sex, and year (1st pass)
  df_tmpc <- foreach (i = 1:nrow(m_geo_age_sex_year), .combine = "rbind", .packages = c("tidyverse", "mgcv")) %dopar% {
    # extract subset of data for region, calculate weekly_death_rate for 100k population
    df_m <- df_summer %>%
      filter(geo == m_geo_age_sex_year$geo[i], age == m_geo_age_sex_year$age[i], sex == m_geo_age_sex_year$sex[i], year == m_geo_age_sex_year$year[i]) %>%
      mutate(weekly_death_rate = 1.0e5 * n_deaths / n_pop)

    # fit individual model with smooth seasonal trend and a smooth term for the mean temperature across the prior 0..3 days, leveraging implicit summation over weekdays
    lag_0_3 <- as.matrix(df_m %>% select(starts_with("tmpc_lag_0_3_dow")))
    gam1 <- bam(weekly_death_rate ~ s(week, k = DOF_WEEK, bs = "ps") + s(lag_0_3, k = DOF_TMPC, bs = "ps"), data = df_m, family = gaussian, discrete = TRUE)

    # interpolate using identical temperature values for all lags and all weekdays (steady state), mid year for calendar weeks
    df_grid <- expand.grid(week = 26, tmpc = seq(round(min(lag_0_3), 1), round(max(lag_0_3), 1), 0.1)) %>% mutate(lag_0_3 = tmpc) %>%
      mutate(response = as.vector(predict.gam(gam1, newdata = ., type = "terms", terms = "s(lag_0_3)")))

    # determine highest temperature for which the smooth temperature response shows a local minimum, and the highest protective/neutral temperature
    cbind(df_grid %>% filter(response <= 0) %>% reframe(t_zero = max(tmpc)),
          df_grid %>% filter(response <= 0) %>% slice_max(order_by = tmpc, n = 2) %>% reframe(s_zero = (response[2] - response[1]) / (tmpc[2] - tmpc[1])),
          df_grid %>% filter(lag(response) >= response, lead(response) >= response) %>% reframe(t_min = max(tmpc)),
          m_geo_age_sex_year[i,], n_pop = unique(df_m$n_pop))
  }

  # document total number of models that were fitted and number of minima and zero crossings found
  df_tmpc %>% group_by(age, sex) %>% reframe(N = n(), ZeroCrossing = sum(t_zero != -Inf), Minima = sum(t_min != -Inf)) %>%
    bind_rows(reframe(., across(everything(), ~ifelse(is.numeric(.), sum(.), "")))) %>%
    select(Age = age, Sex = sex, N, ZeroCrossing, Minima) %>%
    flextable() %>% set_caption("Models: Detected Zero Crossings and Minima") %>% hline(i = 8) %>% autofit()

  # aggregate t_min, t_zero. and s_zero over years, ignoring -Inf (no minimum/no zero crossing found), with a linear trend by country alone
  config_parallel(max_threads = 1)
  lm_min  <- df_tmpc %>% filter(t_min  != -Inf) %>% lm(t_min  ~ 0 + geo:age:sex + country:year, data = ., weights = n_pop)
  lm_zero <- df_tmpc %>% filter(t_zero != -Inf) %>% lm(t_zero ~ 0 + geo:age:sex + country:year, data = ., weights = n_pop)
  lm_sens <- df_tmpc %>% filter(s_zero != -Inf) %>% lm(s_zero ~ 0 + geo:age:sex + country:year, data = ., weights = n_pop)

  # visualize residuals from regression for t_zero
  plot_pdf_png("Figure_X7", aspect_ratio = 0.7,
    df_tmpc %>% filter(t_zero != -Inf) %>% mutate(residuals = lm_zero$residuals) %>%
    ggplot(aes(x = year, y = residuals, group = interaction(age, sex), color = interaction(age, sex))) +
      geom_point(size = 0.2, alpha = 0.5) +
      geom_hline(yintercept = 0.0) +
      labs(color = "Age.Sex", y = "Residuals [° C] from Regression for T_0") +
      scale_color_manual(values = safe_colorblind_palette) +
      std_theme() + theme(legend.position = "right") + facet_wrap(~country))
  print("Figure_X7: Residuals of Regression Fits for T_0 across Years and Regions")
  
  # check for statistically significant slopes for t_zero and s_zero (significant terms country:year)
  tidy(lm_zero, conf.int = TRUE) %>% filter(grepl(":year", term)) %>% arrange(desc(estimate)) %>%
    mutate(Trend = sprintf("%.02f (%.02f to %.02f)", 10.0 * estimate, 10.0 * conf.low, 10.0 * conf.high)) %>%
    mutate(term = str_replace_all(term, c("country" = "", ":year" = ""))) %>%
    select(Country = term, "Trend [C°/10 years]" = Trend) %>%
    flextable() %>% set_caption("Trends in Onset of Heat-Related Risk, by Country") %>% align(align = c("left", "right")) %>% autofit()
  tidy(lm_sens, conf.int = TRUE) %>% filter(grepl(":year", term)) %>% arrange(estimate) %>%
    mutate(Trend = sprintf("%.03f (%.03f to %.03f)", 10.0 * estimate, 10.0 * conf.low, 10.0 * conf.high)) %>%
    mutate(term = str_replace_all(term, c("country" = "", ":year" = ""))) %>%
    select(Country = term, "Trend [1/°C/10 years]" = Trend) %>%
    flextable() %>% set_caption("Trends in Sensitivity at Onset of Heat-Related Risk, by Country") %>% align(align = c("left", "right")) %>% autofit()

  # interpolate regression models and document mean t_zero by country, sex, and age, and by region, sex, and age
  df_tzero_tmin <- m_geo_age_sex_year %>% select(country, geo, age, sex) %>% distinct() %>% mutate(year = LOADS_YEAR) %>%
    mutate(t_zero = predict(lm_zero, newdata = .), t_min = predict(lm_min, newdata = .)) %>% select(!year)
  df_tzero_tmin %>%
    mutate(age_sex = paste(age, sex, sep = ":")) %>%
    group_by(Country = country, age_sex) %>% reframe(t_zero = round(median(t_zero), 1)) %>%   
    pivot_wider(values_from = t_zero, names_from = age_sex) %>%
    bind_rows(reframe(., across(where(is.factor), ~"Mean"), across(where(is.numeric), ~round(mean(.), 1)))) %>%
    flextable() %>%
      set_caption(sprintf("Onset of Heat-Related Risk, for %d, by Country, Sex, and Age Group (Median across Regions, Linear Regression across Years)", LOADS_YEAR)) %>%
      align(align = c("left", rep("center", 8)), part = "all") %>% hline(i = 10) %>% autofit()
  df_tzero_tmin %>%
    mutate(age_sex = paste(age, sex, sep = ":"), t_zero = round(t_zero, 1)) %>% select(!c(country, age, sex, t_min), Region = geo) %>%
    pivot_wider(values_from = t_zero, names_from = age_sex) %>%
    flextable() %>%
      set_caption(sprintf("Onset of Heat-Related Risk, for %d, by Region, Sex, and Age Group (Linear Regression across Years)", LOADS_YEAR)) %>%
      align(align = c("left", rep("center", 8)), part = "all") %>% autofit()

  # visualize onset of temperature risk (zero crossing) on map, by age by sex
  plot_pdf_png("Figure_3", aspect_ratio = 0.6,
    df_tzero_tmin %>%
    left_join(., geo_info, join_by("geo" == "id_region")) %>% st_as_sf() %>%
    ggplot() +
      geom_sf(aes(fill = t_zero)) +
      labs(fill = "T [°C]", y = "© EuroGeographics for the\nadministrative boundaries") +
      scale_fill_gradientn(colors = safe_heatmap_palette, breaks = seq(15, 30, 5), limits = c(13.0, 32.0)) +
      geom_sf(inherit.aes = FALSE, aes(), linewidth = 0.05, color = "black", fill = NA) +
      scale_x_continuous(breaks = seq(-10, 10, 10)) + scale_y_continuous(breaks = seq( 40, 50, 10)) +
      std_theme() + theme(legend.position = "right") + facet_grid(vars(sex), vars(age)))
  print(sprintf("Figure_3: Onset of Heat-Related Risk, for %d, Regression Fit across Years", LOADS_YEAR))

  # visualize onset of temperature risk (zero crossing) on map, by age by sex, compared to max. by region and sex
  plot_pdf_png("Figure_S1", aspect_ratio = 0.6,
    df_tzero_tmin %>% group_by(geo, sex) %>% mutate(t_zero = t_zero - max(t_zero)) %>% ungroup() %>%
    left_join(., geo_info, join_by("geo" == "id_region")) %>% st_as_sf() %>%
    ggplot() +
      geom_sf(aes(fill = t_zero)) +
      labs(fill = "T-Tmax [°C]", y = "© EuroGeographics for the\nadministrative boundaries") +
      scale_fill_gradient(low = "#7F0000", high = "#EFEFEF", breaks = seq(-5, 0, 1)) +
      geom_sf(inherit.aes = FALSE, aes(), linewidth = 0.05, color = "black", fill = NA) +
      scale_x_continuous(breaks = seq(-10, 10, 10)) + scale_y_continuous(breaks = seq( 40, 50, 10)) +
      std_theme() + theme(legend.position = "right") + facet_grid(vars(sex), vars(age)))
  print(sprintf("Figure_S1: Difference of Onset of Heat-Related Risk compared to Max. by Region and Sex, for %d, Regression Fit across Years", LOADS_YEAR))

  # visualize minima in temperature response on map, by age by sex
  plot_pdf_png("Figure_X8", aspect_ratio = 0.6,
    df_tzero_tmin %>%
    left_join(., geo_info, join_by("geo" == "id_region")) %>% st_as_sf() %>%
    ggplot() +
      geom_sf(aes(fill = t_min)) +
      labs(fill = "T [°C]", y = "© EuroGeographics for the\nadministrative boundaries") +
      scale_fill_gradientn(colors = safe_heatmap_palette, breaks = seq(15, 30, 5), limits = c(13.0, 32.0)) +
      geom_sf(inherit.aes = FALSE, aes(), linewidth = 0.05, color = "black", fill = NA) +
      scale_x_continuous(breaks = seq(-10, 10, 10)) + scale_y_continuous(breaks = seq( 40, 50, 10)) +
      std_theme() + theme(legend.position = "right") + facet_grid(vars(sex), vars(age)))
  print(sprintf("Figure_X8: Minima of Heat-Related Risk, for %d, Regression Fit across Years", LOADS_YEAR))

  # shift onset of heat-related risk to 0 °C by region, by age, by sex, save
  df_summer_shifted <- df_summer %>% left_join(., df_tzero_tmin, by = c("country", "geo", "age", "sex")) %>% mutate(across(starts_with("tmpc"), ~.x - t_zero))
  save(df_summer_shifted, file = "../Processed/SummerShifted.Rdata", compress = "xz")
  rm(list = ls(pattern = "lm_sens|lm_zero|lm_min")); stop_gc()
```
# Pass 2 - Common model across regions and years, by age and by sex

Common GAMs are computed for the weekly death rates (assuming a Gaussian error distribution) as a function of the mean temperature across the prior 0..3 days, separately by age groups and by sex, but across regions and years, for observed temperatures and for temperatures relative to the regional reference (onset of heat-related risks).

```{r fit_common_GAMs}
  # fit common models by sex and age group, across regions and years
  config_parallel(max_threads = nrow(m_age_sex))
  observed <- foreach (i = 1:nrow(m_age_sex), .combine = "rbind", .packages = c("tidyverse", "mgcv")) %dopar% {
    # extract subset of data for region, calculate weekly_death_rate for 100k population
    df_m <- df_summer %>% filter(!year %in% COVID_YEARS, age == m_age_sex$age[i], sex == m_age_sex$sex[i]) %>% mutate(weekly_death_rate = 1.0e5 * n_deaths / n_pop)

    # fit common model across all geographies and years, accounting for differences in absolute rate, in secular trends, in recurring seasonal trends,
    # with a common smooth term for the average temperature over the prior 0..3 days with implicit summation over weekdays
    lag_0_3 <- as.matrix(df_m %>% select(starts_with("tmpc_lag_0_3_dow")))
    gam2 <- bam(weekly_death_rate ~ country + s(geo, bs = "sz", by = country) + s(year, bs = "ps", k = DOF_YEAR, by = country) + s(week, k = DOF_WEEK, bs = "ps", by = country) + s(lag_0_3, k = DOF_TMPC, bs = "ps"),
                weights = n_pop, data = df_m, family = gaussian, discrete = TRUE)

    # predictions for steady state, where geo, country, year and week are ignored, but need to be provided
    df_grid <- data.frame(geo = "AT11", country = "Austria", year = max(YEARS_MODEL), week = 26, lag_0_3 = seq(round(min(lag_0_3), 1), round(max(lag_0_3), 1), 0.1))
    predicted_terms <- predict.gam(gam2, newdata = df_grid, type = "terms", terms = "s(lag_0_3)", se.fit = TRUE)

    # response functions, by age and sex, to be aggregated
    cbind(tmpc = df_grid$lag_0_3, age = m_age_sex$age[i], sex = m_age_sex$sex[i], AIC = AIC(gam2),
          as.data.frame(predicted_terms$fit   ) %>% rename_with(~str_replace_all(.x, c("s[(]lag_" = "response_", "[)]" = ""))),
          as.data.frame(predicted_terms$se.fit) %>% rename_with(~str_replace_all(.x, c("s[(]lag_" = "se_", "[)]" = ""))))
  }
  corrected <- foreach (i = 1:nrow(m_age_sex), .combine = "rbind", .packages = c("tidyverse", "mgcv")) %dopar% {
    # extract subset of data for region, calculate weekly_death_rate for 100k population
    df_m <- df_summer_shifted %>% filter(!year %in% COVID_YEARS, age == m_age_sex$age[i], sex == m_age_sex$sex[i]) %>% mutate(weekly_death_rate = 1.0e5 * n_deaths / n_pop)

    # fit common model across all geographies and years, accounting for differences in absolute rate, in secular trends, in recurring seasonal trends,
    # with a common smooth term for the average temperature over the prior 0..3 days with implicit summation over weekdays
    lag_0_3 <- as.matrix(df_m %>% select(starts_with("tmpc_lag_0_3_dow")))
    gam2 <- bam(weekly_death_rate ~ country + s(geo, bs = "sz", by = country) + s(year, bs = "ps", k = DOF_YEAR, by = country) + s(week, k = DOF_WEEK, bs = "ps", by = country) + s(lag_0_3, k = DOF_TMPC, bs = "ps"),
                weights = n_pop, data = df_m, family = gaussian, discrete = TRUE)

    # save subset of data and fitted model for later passes
    save(df_m, gam2, file = sprintf("../Processed/GAM2_%s_%s.Rdata", m_age_sex$age[i], m_age_sex$sex[i]), compress = "xz")

    # predictions for steady state, where geo, country, year and week are ignored, but need to be provided
    df_grid <- data.frame(geo = "AT11", country = "Austria", year = max(YEARS_MODEL), week = 26, lag_0_3 = seq(round(min(lag_0_3), 1), round(max(lag_0_3), 1), 0.1))
    predicted_terms <- predict.gam(gam2, newdata = df_grid, type = "terms", terms = "s(lag_0_3)", se.fit = TRUE)

    # response functions, by age and sex, to be aggregated
    cbind(tmpc = df_grid$lag_0_3, age = m_age_sex$age[i], sex = m_age_sex$sex[i], AIC = AIC(gam2),
          as.data.frame(predicted_terms$fit   ) %>% rename_with(~str_replace_all(.x, c("s[(]lag_" = "response_", "[)]" = ""))),
          as.data.frame(predicted_terms$se.fit) %>% rename_with(~str_replace_all(.x, c("s[(]lag_" = "se_", "[)]" = ""))))
  }
  
  # compares AIC between model
  rbind(observed %>% mutate(model = "Observed"), corrected %>% mutate(model = "Relative")) %>%
    select(Age = age, Sex = sex, model, AIC) %>% distinct() %>% arrange(Age, Sex, model) %>%
    pivot_wider(names_from = model, values_from = AIC, names_prefix = "AIC_") %>% mutate(AIC_Difference = AIC_Observed - AIC_Relative) %>%
    mutate(across(starts_with("AIC_"), ~round(.x, 0)), Sex = ifelse(Sex == "F", "Female", "Male")) %>%
    flextable() %>% set_caption("AIC for common response using observed T and relative T-T0")

  # combine response functions and convert into tidy table
  response_functions <- rbind(observed %>% mutate(model = "Observed"), corrected %>% mutate(model = "Relative")) %>% select(!AIC) %>%
    pivot_longer(!c(tmpc, age, sex, model)) %>%
    mutate(type = gsub("_.*$", "", name), lag = paste(gsub("^[a-zA-Z]*.", "Lag: ", name), "day(s)")) %>% select(!name) %>%
    pivot_wider(names_from = type, values_from = value)

  # visualize heat-related mortality: temperature response
  plot_pdf_png("Figure_4", aspect_ratio = 0.6,
    response_functions %>% arrange(desc(age), sex) %>% mutate(age_sex = fct_inorder(paste(age, sex, sep = ":"))) %>%
      ggplot(aes(x = tmpc, y = response, group_by = age_sex, fill = age_sex, color = age_sex)) +
        geom_ribbon(aes(ymin = response + qnorm(ALPHA * 0.5) * se, ymax = response + qnorm(1.0 - ALPHA * 0.5) * se), alpha = 0.2, color = NA) +
        geom_line() +
        labs(y = "Weekly Mortality Rate [1/100.000], with 95%-CI", x = NULL) +
        scale_color_manual(values = safe_colorblind_palette, name = "Age:Sex") +
        scale_fill_manual(values = safe_colorblind_palette, name = "Age:Sex") +
        facet_wrap(~model, nrow = 1, scales = "free_x", labeller = as_labeller(c("Observed" = "Temperature T [°C]", "Relative" = "Relative Temperature T-T0 [°C]")), strip.position = "bottom") +
        std_theme() + theme(strip.background = element_blank(), strip.placement = "outside"))
  print("Figure_4: Heat-Related Response, Common model, by Sex by Age")

  # compute and report power-law fits
  power_laws <- foreach (i = 1:nrow(m_age_sex), .combine = "rbind", .packages = c("tidyverse", "mgcv")) %dopar% {
    nls1 <- nls(response ~ a + b * tmpc^c, start = list(a = 0.0, b = 0.1, c = 1.5),
      data = response_functions %>% filter(model == "Relative", tmpc > 0.0, age == m_age_sex$age[i], sex == m_age_sex$sex[i]))
    data.frame(power = coef(nls1)["c"], power_cl = confint(nls1, "c", level = 1.0 - ALPHA)[1], power_ch = confint(nls1, "c", level = 1.0 - ALPHA)[2]) %>%
      mutate(Age = m_age_sex$age[i], Sex = m_age_sex$sex[i])
  }
  power_laws %>% mutate(power_ci = sprintf("%.02f (%.02f to %.02f)", power, power_cl, power_ch)) %>%
    select(Age, Sex, Exponent = power_ci) %>% arrange(Age, Sex) %>% mutate(Sex = ifelse(Sex == "F", "Female", "Male")) %>%
    flextable() %>% set_caption("Power Law Approximation to Exposure Response Curves (T >= T_zero)") %>% align(align = "center", part = "all") %>% autofit()
```
# Pass 3 - Load Factors and Trends, GAM followed by LM

The common GAMs from pass 2 for relative temperatures above the onset are re-loaded. Using the common response (by age, by sex) across all countries, regions, and years, the corresponding load factors and their secular trends are computed for each country.

```{r fit_loading_common_models}
  # fit common models by sex and age group, across regions and years
  config_parallel(max_threads = nrow(m_age_sex))
  load_functions <- foreach (i = 1:nrow(m_age_sex), .combine = "rbind", .packages = c("tidyverse", "mgcv", "splines")) %dopar% {
    # reload subset of data and fitted model from Pass 2
    load(file = sprintf("../Processed/GAM2_%s_%s.Rdata", m_age_sex$age[i], m_age_sex$sex[i]))

    # extract matrices for prediction, clipping at 0.0
    lag_0_3         <- as.matrix(df_m %>% select(starts_with("tmpc_lag_0_3_dow")))
    lag_0_3_clipped <- as.matrix(df_m %>% select(starts_with("tmpc_lag_0_3_dow")) %>% mutate(across(everything(), ~pmin(0.0, .))))

    # append response functions, cancelling out the response below the risk onset
    df_m <- df_m %>%
      mutate(t_response = predict.gam(gam2, newdata = list(geo = df_m$geo, country = df_m$country, year = df_m$year, week = df_m$week, lag_0_3 = lag_0_3)) -
                          predict.gam(gam2, newdata = list(geo = df_m$geo, country = df_m$country, year = df_m$year, week = df_m$week, lag_0_3 = lag_0_3_clipped)))

    # extract load functions from a linear model, using the common response as a predictor variable, replacing s(week) by bs(week)
    # and computing country level deviations from the mean response, specified as the offset
    lm3 <- lm(weekly_death_rate ~ geo + country:factor(year) + country:bs(week, df = DOF_WEEK) + offset(t_response) + country:t_response + country:year:t_response,
              weights = n_pop, data = df_m %>% mutate(year = year - LOADS_YEAR))

    # save subset of data and fitted model for later passes
    save(df_m, file = sprintf("../Processed/Common_%s_%s.Rdata", m_age_sex$age[i], m_age_sex$sex[i]), compress = "xz")

    # extract, rename, and return coefficients, i.e., scales and trends
    coefs1 <- as.data.frame(summary(lm3)$coef) %>% mutate(var = rownames(.)) %>% filter(grepl("t_response", var))
    row.names(coefs1) <- NULL
    coefs1 %>% mutate(age = m_age_sex$age[i], sex = m_age_sex$sex[i])
  }

  # extract results from load functions
  results <- load_functions %>% mutate(Type = ifelse(grepl("year", var), "Trend", "Scale")) %>%
    mutate(country = str_replace_all(var, c("country" = "", ":t_response" = "", ":year" = ""))) %>%
    select(Type, country, age, sex, value = Estimate, SE = "Std. Error", p.value = "Pr(>|t|)")

  # document scales and trends
  results %>% arrange(country, age, sex) %>%
    mutate(value = ifelse(Type == "Scale", value + 1.0, value), sex = ifelse(sex == "F", "Female", "Male")) %>%
    mutate(p.value = case_when(p.value < 0.001 ~ "<0.001", p.value < 0.10 ~ sprintf("%.03f", p.value), TRUE ~ sprintf("%.02f", p.value))) %>%
    mutate(value = sprintf("%.02f (%.02f to %.02f),%s", value, value + qnorm(ALPHA * 0.5) * SE, value + qnorm(1.0 - ALPHA * 0.5) * SE, p.value)) %>%
    select(Type, Country = country, Age = age, Sex = sex, value) %>%
    pivot_wider(names_from = Type, values_from = value) %>%
    separate_wider_delim(Scale, names = c("Scale.Value.CI", "Scale.p"), delim = ",") %>%
    separate_wider_delim(Trend, names = c("Trend.Value.CI", "Trend.p"), delim = ",") %>%
    flextable() %>%
      set_caption("Heat Sensitivity: Scale (p for != 1) and Trend (p for != 0) by Country") %>%
      hline(i = seq(8, 72, 8)) %>% align(align = "center", part = "all") %>% align(j = c(1, 4:7), align = "left", part = "all") %>% autofit()

  # aggregate regions to country level
  country_info <- geo_info %>% mutate(country = ifelse(country %in% c("England", "Wales"), "England.Wales", country)) %>%
    group_by(country) %>% reframe(combined_geometry = st_union(geometry))

  # visualize heat sensitivity results
  min_scale <- min(results %>% filter(Type == "Scale") %>% pull("value"))
  max_scale <- max(results %>% filter(Type == "Scale") %>% pull("value"))
  plot_pdf_png("Figure_5", aspect_ratio = 0.6,
    left_join(results %>% filter(Type == "Scale"), country_info, by = "country") %>% st_as_sf() %>%
    ggplot() +
      geom_sf(aes(geometry = combined_geometry, fill = value + 1.0)) +
      scale_fill_gradientn(values = c(0.0, min_scale / (min_scale - max_scale), 1.0), colours = c("#007F00", "#FFFFFF", "#FF0000")) +
      labs(fill = "Scale", x = "", y = "© EuroGeographics for the\nadministrative boundaries") +
      scale_x_continuous(breaks = seq(-10, 10, 10)) + scale_y_continuous(breaks = seq( 40, 50, 10)) +
      std_theme() + theme(legend.position = "right", legend.text = element_text(hjust = 1)) + facet_grid(vars(sex), vars(age)))
  print(sprintf("Figure_5: Heat Sensitivity (%d): Scale x Common Age-Sex Response (across Countries and %d-%d)", LOADS_YEAR, min(YEARS_MODEL), max(YEARS_MODEL)))

  # visualize heat sensitivity trend results
  min_trend <- min(results %>% filter(Type == "Trend") %>% pull("value"))
  max_trend <- max(results %>% filter(Type == "Trend") %>% pull("value"))
  plot_pdf_png("Figure_6", aspect_ratio = 0.6,
    left_join(results %>% filter(Type == "Trend"), country_info, by = "country") %>% st_as_sf() %>%
    ggplot() +
      geom_sf(aes(geometry = combined_geometry, fill = value)) +
      scale_fill_gradientn(values = c(0.0, min_trend / (min_trend - max_trend), 1.0), colours = c("#007F00", "#FFFFFF", "#FF0000")) +
      labs(fill = "Annual\nScale\nChange\n[1/year]", x = "", y = "© EuroGeographics for the\nadministrative boundaries") +
      scale_x_continuous(breaks = seq(-10, 10, 10)) + scale_y_continuous(breaks = seq( 40, 50, 10)) +
      std_theme() + theme(legend.position = "right", legend.text = element_text(hjust = 1)) + facet_grid(vars(sex), vars(age)))
  print(sprintf("Figure_6: Heat Sensitivity Trend (%d-%d): Slope = Annual Change to Factor", min(YEARS_MODEL), max(YEARS_MODEL)))
```
# Pass 4 - Heat-Related and Excess Mortality, Calibration Plot

Exploratory calculations to assess plausibility of the model: Table of juxtaposed heat-related and excess mortality by country by year, summed over regions, sex, and age groups, and calibration plot between observed and predicted mortality for all weeks, by country.

```{r heat_related_and_excess_mortality_calibration}
  # fit common models by sex and age group, across regions and years
  config_parallel(max_threads = nrow(m_age_sex))
  predicted_mortality <- foreach (i = 1:nrow(m_age_sex), .combine = "rbind", .packages = c("tidyverse", "mgcv", "splines")) %dopar% {
    # reload subset of data with common response
    load(file = sprintf("../Processed/Common_%s_%s.Rdata", m_age_sex$age[i], m_age_sex$sex[i]))

    # determine flexible load functions, by country and by year
    lm4 <- lm(weekly_death_rate ~ geo + country:factor(year) + country:bs(week, df = DOF_WEEK) + offset(t_response) + country:factor(year):t_response,
              weights = n_pop, data = df_m)
    
    # compute heat-related mortality as the difference between the predicted mortality for observed temperatures and the predicted mortality at onset
    # compute excess mortality as the difference between the observed mortality and the predicted mortality at onset
    df_m <- df_m %>%
      mutate(predicted_deaths       = round(n_pop * 1.0e-5 * predict(lm4, type = "response"                                             ), 0),
             predicted_deaths_onset = round(n_pop * 1.0e-5 * predict(lm4, type = "response", newdata = df_m %>% mutate(t_response = 0.0)), 0))
  }

  # create a replica of Figure 2, for predicted mortality instead of observed mortality
  incremental_death_rate_by_country_exposure_by_region <- predicted_mortality %>%
    mutate(tmpc_max_lag_0_3 = round(pmax(tmpc_lag_0_3_dow_0, tmpc_lag_0_3_dow_1, tmpc_lag_0_3_dow_2, tmpc_lag_0_3_dow_3,
                                         tmpc_lag_0_3_dow_4, tmpc_lag_0_3_dow_5, tmpc_lag_0_3_dow_6), 0)) %>%
    group_by(geo, year, sex, age) %>% mutate(n_deaths = predicted_deaths - mean(predicted_deaths)) %>% ungroup() %>%
    mutate(decade = sprintf("%d-%d", 10 * (year %/% 10), 10 * (year %/% 10) + 9)) %>%
    group_by(country, decade, age, tmpc_max_lag_0_3) %>% reframe(r_deaths = 100000.0 * sum(n_deaths) / sum(n_pop), n_pop = sum(n_pop))
  plot_pdf_png("Figure_S4", aspect_ratio = 0.7,
    incremental_death_rate_by_country_exposure_by_region %>%
    filter(age %in% c("75-84", "85-99"), decade %in% c("2000-2009", "2010-2019")) %>%
    ggplot(aes(x = tmpc_max_lag_0_3, y = r_deaths, group_by = interaction(age, decade), color = age, weight = n_pop)) +
      geom_smooth(aes(fill = age, color = age, linetype = decade), formula = y ~ x, method = "loess", span = 0.8, se = TRUE, alpha = 0.2) +
      labs(color = "", fill = "", linetype = "",
           x = "Mean Temperature over prior 0:3 Days (relative to T0), Maximum over 7 Days of Week",
           y = "Weekly Rate of Incremental Deaths in Summer [1/100k]\nvs. Region-Year-Sex-Age Mean, All Regions") +
      scale_x_continuous(breaks = seq(-10, 10, 10), labels = ~sprintf("%d °C", .x)) +
      scale_fill_manual(values = safe_colorblind_palette[c(4,2)]) +
      scale_color_manual(values = safe_colorblind_palette[c(4,2)]) +
      guides(fill = guide_legend(nrow = 1, byrow = FALSE), linetype = guide_legend(nrow = 1, byrow = FALSE, override.aes = list(color = "black", fill = "#BFBFBF"))) +
      std_theme() + theme(legend.position = "top") + facet_wrap(~country, nrow = 2))
  print("Figure_S4: Modeled Exposure Response Curves (Summer), Incremental Deaths, 74-84 and 85+ Years")

  # create visual comparison for heat waves, 2003, one country
  plot_pdf_png("Figure_S5", aspect_ratio = 0.7,
    predicted_mortality %>% filter(country == "Germany", year == 2003) %>%
      group_by(sex, age, week) %>% reframe(across(c(n_deaths, predicted_deaths, predicted_deaths_onset), ~sum(.x))) %>%
    ggplot(aes(x = week)) +
      geom_line(aes(y = n_deaths), color = "black") +
      geom_line(aes(y = predicted_deaths), color = "red") +
      geom_line(aes(y = predicted_deaths_onset), color = "blue") +
      expand_limits(y = 0.0) +
      labs(x = "Week of 2003", y = "Number of Deaths") +
      facet_grid(vars(sex), vars(age), scales = "free"))

  # compute heat-related mortality as the difference between the predicted mortality for observed temperatures and the predicted mortality at the onset
  # compute excess mortality as the difference between the observed mortality and the predicted mortality at the onset, in heat wave weeks
  heat_related_excess_mortality <- predicted_mortality %>% filter(geo != "UKL") %>%
    mutate(heat_related_deaths = predicted_deaths - predicted_deaths_onset, excess_deaths = (n_deaths - predicted_deaths_onset) * (predicted_deaths != predicted_deaths_onset)) %>%
    group_by(country, year, age, sex) %>% reframe(across(c(heat_related_deaths, excess_deaths), ~sum(.x)))

  # compute Pearson correlation by country
  heat_related_excess_mortality %>%
    group_by(Country = country, Year = year) %>% reframe(HR = sum(heat_related_deaths), EX = sum(excess_deaths)) %>%
    group_by(Country)                        %>% reframe(rho = cor(x = HR, y = EX, method = "pearson")) %>%
    mutate(Country = gsub(".Wales", "", Country)) %>% arrange(desc(rho)) %>% mutate(rho = sprintf("%.03f", rho)) %>%
    flextable() %>% set_caption("Pearson correlation between Heat-Related and Excess Mortality") %>% autofit()

  # create table of heat-related and excess mortality
  heat_related_excess_mortality %>%
    group_by(Country = country, Year = year) %>% reframe(HR_EX = sprintf("%.0f/%.0f", sum(heat_related_deaths), sum(excess_deaths))) %>%
    mutate(Country = gsub(".Wales", "", Country), Year = as.character(Year)) %>%
    pivot_wider(values_from = HR_EX, names_from = Country) %>%
      flextable() %>% set_caption("Heat-Related/Excess Deaths by Country and Year") %>% align(align = "center", part = "all")

  # check predicted vs. observed mortality, aggregated by country
  calibration <- predicted_mortality %>%
    group_by(country, sex, age, year, week) %>% reframe(obs_mort = sum(n_deaths) * 1.0e5 / sum(n_pop), pre_mort = sum(predicted_deaths) * 1.0e5 / sum(n_pop))
  plot_pdf_png("Figure_S2", aspect_ratio = 0.7,
    calibration %>% filter(obs_mort > 0, pre_mort > 0) %>%
      ggplot(aes(x = obs_mort, y = pre_mort, color = age)) +
        geom_abline(slope = 1.0, intercept = 0.0) + geom_point(size = 0.5) + 
        scale_x_continuous(trans = "log10") + scale_y_continuous(trans = "log10") +
        scale_color_manual(values = safe_colorblind_palette) +
        labs(x = "Observed Weekly Mortality Rate [1/100k]", y = "Predicted Weekly Mortality Rate [1/100k]", color = "Age") +
        std_theme() + theme(legend.position = "top") +
        facet_wrap(~country, nrow = 2))
  print("Figure_S2: Model Calibration Plot")
```
# Done

```{r done}
  print(proc.time() - start_time); rstudioapi::versionInfo(); sessionInfo()
```